---
title: 'Course 6 - Introduction to Data Visualization with ggplot2'
author: "Karim Badr"
date: "06/03/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Let's read all data files first.

```{r}
load("/Users/karimbadr/Documents/GitHub/DataCamp/Data Scientist with R/Notes/iris.RData")
load("/Users/karimbadr/Documents/GitHub/DataCamp/Data Scientist with R/Notes/diamonds.RData")
load("/Users/karimbadr/Documents/GitHub/DataCamp/Data Scientist with R/Notes/fish.RData")
load("/Users/karimbadr/Documents/GitHub/DataCamp/Data Scientist with R/Notes/recess.RData")
```

```{r}
library(ggplot2)
```

# Chapter 1: Introduction

Exploratory vs Explanatory visualizations 

Exploratory visualizations are easily-generated, data-heavy and intended for a small specialist audience, for example yourself and your colleagues - their primary purpose is graphical data analysis. Explanatory visualizations are labor-intensive, data-specific and intended for a broader audience, e.g. in publications or presentations - they are part of the communications process. 

Anscombe's plots : Anscombe's quartet comprises four data sets that have nearly identical simple descriptive statistics, yet have very different distributions and appear very different when graphed. Each dataset consists of eleven points.

```{r}
# Change the command below so that cyl is treated as factor
ggplot(mtcars, aes(factor(cyl), mpg)) +
  geom_point()
```

The grammar of graphics (Leland Wilkinson):

2 principles: 
1) Graphics = distinct layers of grammatical elements
2) Meaningful plots through aesthetic mappings

Layers are like adjectives and nouns and aesthetic mappings are like grammatical rules for how to assemble that vocabulary.

3 essential grammatical elements: data, aesthetics, and geometries 

Aesthetics: the scales onto which we map our data
Geometries: the visual elements used for our data
Themes: all non-data ink

7 grammatical elements (or layers) to be explored: data, aesthetics, geometries, themes, statistics, coordinates, and facets.

With aes : color, size, and shape.
Shape is only used with categorical data.

Two common geom layer functions:

* geom_point() adds points (as in a scatter plot).
* geom_smooth() adds a smooth trend curve.

If you have multiple geoms, then mapping an aesthetic to data variable inside the call to ggplot() will change all the geoms. It is also possible to make changes to individual geoms by passing arguments to the geom_*() functions.

You can save a ggplot to a variable and then add onto that.

You can use aes() also inside a geom.

# Chapter 2: Aesthetics

X and Y axes are aesthetics.
Color aesthetic vs Fill color : color is usually color of points or borders (outlines) of other geoms.
Aesthetics are called upon inside aes().
You can use aes() also inside a geom. We do that if we are mixing data sources or if we dont want all layers to inherit the same aesthetics.
Size : area
alpha: transperancy
linetype: line dash pattern
labels (text on plot or axis)
shape : shape of a point 

These are the aesthetics you can consider within aes() in this chapter: x, y, color, fill, size, alpha, labels and shape.

You used aes() inside a geom usually when you are adding to a ggplot that was saved as a variable [usually not always!]

label and shape are only applicable to categorical data.

When we talk about how something looks, we refer to it as attributes. Attributes are set in the geom arguments!

Modifying aesthetics: Positions (for overlapping) : identity (default), dodge, stack,fill,jitter,jitterdodge, and nudge

jitter adds random noise to distinguish areas of high density. you can define it within geom_point() or as its own function first as jitter_position() and then use that in the geom_point(). 

can also use geom_jitter() instead of geom_point(position="jitter")

Many scale functions also can be used : scale_*_*

The arguments for the scale functions are usually the name of the scale, limits, breaks, expand and labels. 

Limits: scale's range
Breaks: Tick mark positions
Expand: expand is a numeric vector of length two, giving a multiplicative and additive constant used to expand the range of the scales so that there is a small gap between the data and the axes.
Labels: adjust category names

we can use labs() to quickly change the axes labels (x,y, and legend [e.g color])

scale_color_manual() defines properties of the color scale (i.e. axis). The first argument sets the legend title. values is a named vector of colors to use.

geom_bar(position="dodge") : bars are displayed next to each other instead on top of each other

Best practices: FORM FOLLOWS FUNCTION

Typically, the dependent variable is mapped onto the the y-axis and the independent variable is mapped onto the x-axis.

# Chapter 3: Geometries

1st layer: data
2nd layer: aesthetics
3rd layer: geometries


1) Scatter Plots

geom_point() : you need an x and y aesthetic
Optional: alpha, color, fill, size, shape, and stroke
geom_point() can inherit data from a different dataset!

Alpha or hollow circles help with overplotting

Possible geoms for scatter plots: points, jitter, abline, smooth, and count

2) Histograms

A histogram is a special case of a bar plot.

we show the distribution of a continuous variable

here we only need a single aesthetic x

The default binwidth is calculated as follows:

```{r}
diff(range(iris$Sepal.Width))/30
```

X-axis labels should be between the bars not exactly on them because they represnet intervals. to do that: geom_histogram(bindwidth=t,center=t/2)

The default position of a histogram is position="stack". You can use dodge or fill instead. 

In aes, Map y to the internal variable ..density.. to show frequency densities.

  geom_histogram(binwidth = 1, position = "identity",alpha=0.4) : bars are top of each other, using the "identity" position. So each bar can be seen, set alpha to 0.4.
  
3) Bar Plots

Bar plots used with categorical x axis
You can use geom_bar() or geom_col()
geom_car() : stat is count : it counts the actual number of cases at each x position
geom_col() : stat is identity : plots actual values

Two types: absolute counts and distributions

geom_col() is just geom_bar() where both the position and stat arguments are set to "identity"

aes(fill=z) is usually used to specify colors

scale_fill_brewer() can be used. this, without arguments, has max of 9 colors


4) Line Plots

Line plots : line or path

useful for time series

geom_area()

geom_ribbon() for overlapping areas with alpha

specifying color in aes() is important for many groups of lines.

# Chapter 4: Themes

3 types: text, line, and rectangle.

modified using: element_text(), element_line(), or element_rect()

To modify an element, just call its argument in the theme function and use the appropriate element_ function to specify what we want to change

Hierarchical naming reflects inheritance rules: All text elements inherit from text, so if we changed that argument, all downstream arguments would be affected.

To remove any item, use element_blank().

You usually define what element_*() you're using

For the following u dont need that:

Whitespace means all the non-visible margins and spacing in the plot.

To set a single whitespace value, use unit(x, unit), where x is the amount and unit is the unit of measure.

Borders require you to set 4 positions, so use margin(top, right, bottom, left, unit). To remember the margin order, think TRouBLe.

The default unit is "pt" (points), which scales well with text. Other options include "cm", "in" (inches) and "lines" (of text).


EG: axis.tick.length(), legend.key.size(),legend.margin(), plot.margin()


Theme flexibility:

We can create our own themes!

You can create a theme() object and save it and then add that to your plot!

built in theme functions from ggplot or ggthemes packages

update theme to default: original<-theme_update() then theme_set(original)

Full example:

```{r}
# Set the color scale
palette <- brewer.pal(5, "RdYlBu")[-(2:4)]

# Add a title and caption
ggplot(gm2007, aes(x = lifeExp, y = country, color = lifeExp)) +
  geom_point(size = 4) +
  geom_segment(aes(xend = 30, yend = country), size = 2) +
  geom_text(aes(label = round(lifeExp,1)), color = "white", size = 1.5) +
  scale_x_continuous("", expand = c(0,0), limits = c(30,90), position = "top") +
  scale_color_gradientn(colors = palette) +
  labs(title="Highest and lowest life expectancies, 2007",
  caption="Source: gapminder")+
  theme_classic() +
  theme(axis.line.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text = element_text(color="black"),
        axis.title = element_blank(),
        legend.position = "none")+
    geom_vline(xintercept=global_mean, color="grey40", linetype=3)+
    annotate(
    "text",
    x = x_start, y = y_start,
    label = "The\nglobal\naverage",
    vjust = 1, size = 3, color = "grey40"
  )+
  annotate(
    "curve",
    x = x_start, y = y_start,
    xend = x_end, yend = y_end,
    arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
    color = "grey40"
  )
```

