---
title: 'Course 5 - Joining Data with dplyr'
author: "Karim Badr"
date: "22/2/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Chapter 1: Joining Tables

```{r}
library(dplyr)
library(tidyr)
```

```{r}
#file.choose() #to choose file
```

```{r}
sets<-readRDS("/Users/karimbadr/Desktop/Karim/Learning/DataCamp/R Data Scientist/sets.rds")
themes<-readRDS("/Users/karimbadr/Desktop/Karim/Learning/DataCamp/R Data Scientist/themes.rds")
```

Inner Join

```{r}
sets %>% 
  inner_join(themes,by=c("theme_id"="id"),suffix=c("_set","_theme"))
```

One-to-many relationships

```{r}
inventories<-readRDS("/Users/karimbadr/Desktop/Karim/Learning/DataCamp/R Data Scientist/inventories.rds")
```

```{r}
sets %>% 
  inner_join(inventories,by=c("set_num"))
```
```{r}
sets %>% 
  inner_join(inventories,by=c("set_num")) %>% 
  filter(version==1)
```
Joining in either direction will give the same thing!

We can use the same approach to join 3 or more tables with a series of joins.

# Chapter 2: Left and Right Joins

LEFT JOIN verb

Inner join only keeps observations that appear in both tables. Left join keeps all from 1st and matches to second.

```{r}
inventory_parts<-readRDS("/Users/karimbadr/Desktop/Karim/Learning/DataCamp/R Data Scientist/inventory_parts.rds")
```

```{r}
inventory_parts_joined<-inventories %>% 
  inner_join(inventory_parts,by=c("id"="inventory_id")) %>% 
  select(-id,-version) %>% 
  arrange(desc(quantity))
```

```{r}
batmobile<-inventory_parts_joined %>% 
  filter(set_num=="7784-1") %>% 
  select(-set_num)

batwing<-inventory_parts_joined %>% 
  filter(set_num=="70916-1") %>% 
  select(-set_num)
```

```{r}
batmobile %>% 
  left_join(batwing,by=c("part_num","color_id"),suffix=c("_batmobile","_batwing"))
```
RIGHT JOIN verb

Right join keeps all in 2nd.

```{r}
batmobile %>% 
  right_join(batwing,by=c("part_num","color_id"),suffix=c("_batmobile","_batwing"))
```
Left joins and right joins are mirrors of each others.

```{r}
#replace_na(list(n=0))  n is the name of the column
```


Joining a table onto itself when there is a hierarchical relationship (e.g. column parent_id)

```{r}
themes %>% 
  inner_join(themes,by=c("parent_id"="id"),suffix=c("_child","_parents"))
```

# Chapter 3: Full, Semi, and Anti Joins


FULL Join : keeping all observations, whether they appear in one or both

```{r}
batmobile$part_num<-as.numeric(batmobile$part_num)
```

```{r}
batmobile %>% 
  full_join(batwing,by=c("part_num"="color_id"),suffix=c("_batmobile","_batwing")) %>% 
  replace_na(list(quantity_batmobile=0,
                  quantity_batwing=0))
```
The 4 verbs: left,right,inner, and full are mutating verbs.

There's a different class of verbs called filtering joins, which has SEMI and ANTI joins.
They keep or remove observations from the first table but doesn't add new variables. 

SEMI Join: What observations in X are also in Y but returns only X


ANTI Join: What observations in X are not in Y

# Chapter 4: Case Study: Joins on Stack Overflow Data

To count within a summarize, use n(). 

The bind_rows() verb : combine rows from different tables (stack one on top of another)

year() function from lubridate package