---
title: 'Course 4: Data Manipulation with dplyr'
author: "Karim Badr"
date: "21/2/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Chapter 1: Tranforming data with dplyr

```{r}
library(dplyr)
```

Verbs: select(), filter(), arrange(), and mutate()

```{r}
counties<-readRDS("/Users/karimbadr/Desktop/Karim/Learning/DataCamp/R Data Scientist/counties.rds")
```

```{r}
glimpse(counties)
```

```{r}
counties_selected<-counties %>% 
  select(state,county,population,unemployment)
```

```{r}
counties_selected %>% 
  arrange(desc(population)) %>% 
  filter(state=="New York",unemployment<6)
```
```{r}
counties_selected %>% 
  mutate(unemployed_population=population*unemployment/100)
```

# Chapter 2: Aggregating data

COUNT verb

```{r}
counties %>% count()
```
```{r}
counties %>% count(state,sort=TRUE)
```

You can also add a weight

```{r}
counties %>% 
  count(state,wt=population,sort=TRUE)
```
GROUP

```{r}
counties %>% 
  group_by(state,metro) %>% 
  summarise(total_population=sum(population),average_unemployment=mean(unemployment)) %>% 
  arrange(desc(average_unemployment)) %>% 
  ungroup()
```

TOP_N

```{r}
counties %>% 
  group_by(state) %>% 
  top_n(1,population)
```

# Chapter 3: Selecting and transforming data

SELECT
```{r}
counties %>% 
  select(state,county,drive:work_at_home)
```
```{r}
counties %>% 
  select(state,county,contains("work"))
```

```{r}
counties %>% 
  select(state,county,starts_with("income"))
```
```{r}
?select_helpers
```

```{r}
counties %>% 
  select(-census_id)
```
RENAME verb

```{r}
counties_selected %>% 
  rename(unemployment_rate=unemployment)
```
```{r}
counties %>% 
  select(state,county,population,unemployment_rate=unemployment)
```
TRANSMUTE verb

combination of select and mutate

```{r}
counties %>% 
  transmute(state,county,fraction_men=men/population)
```

# Chapter 4: Case study: The babynames dataset

```{r}
%in%
```

Grouped mutate

```{r}
group_by() %>% 
  mutate() %>% 
  ungroup()
```

Window function: takes a vector and returns another vector of the same length

```{r}
v<-c(1,3,6,14)
lag(v)
```

```{r}
v-lag(v)  #consecutive change
```

