---
title: 'Course 7 - Intermediate Data Visualization with ggplot2'
author: "Karim Badr"
date: "06/03/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Chapter 1: Statistics

Either called from within a geom or independently. 

geom_bar <-> stat_count

geom_histogram / geom_freqpoly <-> stat_bin

geom_smooth <-> stat_smooth

geom_smooth uses method=LOESS by default : a non-parametric smoothing algorithm when <1000 obs. Uses weighted mean.
It passes a sliding window across x-axis, controlled by the span argument. smaller span --> more noise
can use different methods such as lm or other parametric methods. Groups >1000's default is gam.

For parametric models, you can use the fullrange=TRUE argument to show the entire range of predictions. 

If you set se=FALSE, you dont get the prediction gray area (se ribbon).

```{r}
stat_smooth(mapping=aes(group=1),method="lm",se=FALSE)
```

this does the smoothing for all values even though we have that done seperately for each group before.

LOESS isn't great on very short sections of data; better to use lm for short sections and LOESS for overall. The default span for LOESS is 0.9

4 causes for over-plotting: larged datasets, aligned values on a single axis, low-precision data, and integer data

the last 2 causes can be dealt with using geom_count(). geom_count() is associated with stat_sum()

geom_quantile() is robust as opposed to linear models with means that are not robust
you would use quantile regression when you have heteroscedasticity (variance across predictor variable is not consistent). geom_quantile() is associated with stat_quantile().

Stats: from library(Hmisc), use smean.sdl(x,mult=1) to find the mean and plus or minus 1 SD (thats why mult=1). the equivalent to this in using it in ggplot2 is mean_sdl(x,mult=1)

```{r}
ggplot(iris,aes(x=Species,y=Sepal.Length))+
  stat_summary(fun.data=mean_sdl,fun.args=list(mult=1))
```

```{r}
#error bars
ggplot(iris,aes(x=Species,y=Sepal.Length))+
  stat_summary(fun.y=mean,geom="point")+
  stat_summary(fun.data=mean_sdl,fun.args=list(mult=1),geom="errorbar",width=0.1)
```

qt(p,df) gives you the t score using a degrees of freedom df and p which is the area to the left so for 95% p=0.975

can also use from Hmisc this: mean_cl_normal(x)

IMPORTANT:

stat_summary(): summarize y values at distinct x values.
stat_function(): compute y values from a function of x values
stat_qq(): perform calculations for a quantile-quantile plot [QQ Plot].

```{r}
mam.new <- data.frame(body = log10(iris$Sepal.Length))

ggplot(mam.new, aes(x = body)) +   
  geom_histogram(aes( y = ..density..)) +  
  geom_rug() +  
  stat_function(fun = dnorm, color = "red",
                args = list(mean = mean(mam.new$body),sd = sd(mam.new$body)))
```

```{r}
ggplot(mam.new, aes(sample = body)) +
  stat_qq() +  
  geom_qq_line(col = "red")

```

# Chapter 2: Coordinates

# Chapter 3: Facets

# Chapter 4: Best Practices