---
title: 'Course 3: Introduction to the Tidyverse'
author: "Karim Badr"
date: "21/2/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Chapter 1: Data Wrangling

```{r}
library(tidyverse)
```

```{r}
install.packages("gapminder")
library(gapminder)
```

```{r}
gapminder
```
Useful verbs:

1) FILTER
```{r}
gapminder %>% 
  filter(year==2007)
```
```{r}
gapminder %>% 
  filter(year==2007,country=="United States")
```

2) ARRANGE

```{r}
gapminder %>% 
  arrange(gdpPercap)
```
```{r}
gapminder %>% 
  arrange(desc(gdpPercap))
```
3) MUTATE

```{r}
gapminder %>% 
  mutate(pop=pop/1000000)
```

You can combine all above verbs.

# Chapter 2: Data Visualization

```{r}
gapminder_2007<-gapminder %>% 
  filter(year==2007)
```

```{r}
library(ggplot2)
ggplot(gapminder_2007,aes(x=gdpPercap,y=lifeExp))+
  geom_point()
```

Log scales

We see that most data points in the left are crammed next to each other. The scale spans several orders of magnitude, so it is useful to use log scales.

```{r}
ggplot(gapminder_2007,aes(x=gdpPercap,y=lifeExp))+
  geom_point()+
  scale_x_log10()
```
x and y are aesthetics
Two more aesthetics: color and size
cateogrical-->color
continuous-->size
```{r}
ggplot(gapminder_2007,aes(x=gdpPercap,y=lifeExp,color=continent,size=pop))+
  geom_point()+
  scale_x_log10()
```

Faceting:

Divide your plot into subplots.

```{r}
ggplot(gapminder_2007,aes(x=gdpPercap,y=lifeExp))+
  geom_point()+
  scale_x_log10()+
  facet_wrap(~continent)
```


# Chapter 3: Grouping and Summarizing

SUMMARIZE verb

```{r}
gapminder %>% 
  summarize(meanLifeExp=mean(lifeExp))
```

GROUP_BY verb

```{r}
gapminder %>% 
  group_by(year) %>% 
  summarise(meanLifeExp=mean(lifeExp),totalPop=sum(pop))
```

Visualizing summarised data

```{r}
by_year<-gapminder %>% 
  group_by(year) %>% 
  summarise(meanLifeExp=mean(lifeExp),totalPop=sum(pop))
```

```{r}
ggplot(by_year,aes(x=year,y=totalPop))+
  geom_point()+
  expand_limits(y=0)
```

# Chapter 4: Types of Visualizations

1) LINE PLOTS: useful for showing change over time

```{r}
ggplot(by_year,aes(x=year,y=totalPop))+
  geom_line()+
  expand_limits(y=0)
```

2) BAR PLOTS: comparing statistics for each of several categories (discrete)


```{r}
by_cotninent<-gapminder %>% 
  filter(year==2007) %>% 
  group_by(continent) %>% 
  summarise(meanLifeExp=mean(lifeExp))
```

```{r}
ggplot(by_cotninent,aes(x=continent,y=meanLifeExp))+
  geom_bar(stat="identity")
```

```{r}
ggplot(by_cotninent,aes(x=continent,y=meanLifeExp))+
  geom_col()
```


3) HISTOGRAM: describe the distribution of a one-dimensional numeric variable

```{r}
ggplot(gapminder_2007,aes(x=lifeExp))+
  geom_histogram(binwidth = 5)
```

```{r}
ggplot(gapminder_2007,aes(x=lifeExp))+
  geom_histogram(bins=5)
```

4) BOX PLOTS: compare the distribution of a numeric variable among several categories

```{r}
ggplot(gapminder_2007,aes(x=continent,y=lifeExp))+
  geom_boxplot()+
  ggtitle("Boxplot")
```

